Decoding Mortality: Trends and Causes in Canada between 2001-2022

### 2024-03-17 outputs/paper/paper.qmd
Updated the paper and references.bib

---
title: "Decoding Mortality: Trends and Causes in Canada between 2001-2022"
subtitle: "A Bayesian Statistical Analysis Unveiling the Predominance of Malignant Neoplasms and Heart Diseases"
author: 
  - Yuchao Niu
thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
date: today
date-format: long
abstract: "The study of mortality offers key public health insights into the fundamental causes and trends of death. This research investigates Canada's mortality trends from 2001 to 2022, employing Bayesian Poisson and negative binomial regression models to analyze data from the Canadian Vital Statistics Death Database. It reveals that malignant neoplasms and heart diseases significantly elevate mortality rates, surpassing the effects of unintentional accidents by five times and three times, respectively. Such findings highlight the critical necessity for targeted public health initiatives focusing on the prevention, early detection, and treatment of these conditions."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)
library(arrow)

@@

The original dataset obtained from the Statistics Canada open government portal includes 2,268 observations across 18 variables, offering detailed demographic and geographical information on registered deaths. I aim to focus on five specific variables: 'Reference Period', 'Leading Causes of Death (ICD-10)', 'Characteristics', 'Ranking', and 'Number of Deaths'. The 'Reference Period' denotes the year of death registration (2000-2022), while 'Leading Causes of Death' categorizes the cause according to the International Statistical Classification of Diseases and Related Health Problems (ICD-10). 'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the disease's rank in a given year, respectively. A sample of the cleaned data can be found in @sec-data-details.

@@

R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data downloading, cleaning and visualization.

```{r}

cleaned_data <- read.csv("../data/cleaned_data.csv")

@@

```
```{r}
#| label: fig-1
#| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
#| echo: false
#| warning: false
#| message: false

top_five <-
  transformed_data |>
  filter(
    Year == 2022, 
    n == 22
  ) |>
  slice_max(order_by = desc(Ranking), n = 5) |>
  pull(Cause)

# analysis_data <- cleaned_data |>
#   filter(Cause %in% canada_top_five)

cleaned_data |>
  filter(Cause %in% top_five)|>
  ggplot(aes(x = Year, y = Death, color = Cause)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Year", y = "Annual number of deaths in Canada") +
  facet_wrap(vars(Cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")

@@

```{r}
#| label: fig-2
#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
#| warning: false
#| message: false

cleaned_data |>
  filter(Year == 2021)
  ggplot(top_five, aes(x="", y=total_deaths, fill=cause_of_death)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(fill="Cause of Death")

@@

\begin{align} 
y_i&\sim \text{Poisson}(\lambda_i) \quad \text{(1)} \\
\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
\end{align}

@@

\begin{align} 
y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad \text{(1)} \\
\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
\phi &\sim \text{Gamma}(2, 0.1) \quad \text{5} \\
\end{align}

@@

```{r}
#| label: tbl-2
#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
#| echo: false

@@

@tbl-3 presents coefficients from both Poisson and Negative Binomial regression models for various causes of death. The intercept, representing the baseline of Accidents (unintentional injuries), indicates the log count of deaths when no other specific cause is considered. Coefficients for causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of the heart, and malignant neoplasms reflect log-relative differences in death counts relative to Accidents. Positive coefficients indicate a positive association with the number of deaths.

```{r}
#| label: tbl-3
#| fig-cap: Modeling the most prevalent cause of deaths in Canada, 2001-2022
#| echo: false
#| warning: false
#| message: false

@@

coef_short_names <- c(
  "(Intercept)" = "(Intercept)",
  "CauseCerebrovascular diseases " = "Cerebrovascular diseases",
  "CauseChronic lower respiratory diseases " = "Chronic lower respiratory diseases",
  "CauseDiseases of heart " = "Diseases of heart",
  "CauseMalignant neoplasms " = "Malignant neoplasms"
)

modelsummary(
  list(
    "Poisson" = canada_death_poisson,
    "Negative binomial" = canada_death_negbinomial
  ),
  coef_map = coef_short_names,
  statistic = "std.error"
)

@@

Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts. The Negative Binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794. The negative binomial model's superior fit is further supported by a higher ELPD value compared to the Poisson model. Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.

```{r}

@@

loo_compare(poisson, neg_binomial)

@@

While the Poisson and Negative Binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors. Additionally, we only considered one variable the leading cause of death, without considering rarer causes or potential confounders like socioeconomic factors. This constrains the ability of the model to capture the nuances of mortality and its drivers.

@@

```{r}
#| label: tbl-5
#| fig-cap: A sample of the cleaned dataset, 2001-2022
#| echo: false
#| warning: false
#| message: false
sampled_data <- sample_n(cleaned_data, 5) %>%
  kable()
sampled_data
```

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check for the Poisson regression model described in @sec-4. This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align reasonably well with the actual data (represented by the dark blue line) with minor misalignment. This suggests the model represent the data well.

In @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the negative binomial regression model described in @sec-4. This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align generally well with the actual data (represented by the dark blue line). This suggests the model represents the data well.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
#| fig-subcap: ["Poisson model", "Negative binomial model"]

@@

## Diagnostics
```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
#| fig-subcap: ["Poisson model", "Negative binomial model"]

@fig-stanareyouokay-1 is a trace plot for the Poisson regression model. It shows a horizontal, dense band of samples without any systematic patterns, drifts, or long periods of stagnation. This pattern suggests that the chain is mixing well and sampling efficiently from the posterior distribution. This suggests the Poisson regression model is suitable.

@fig-stanareyouokay-2 is a Rhat plot. It shows an Rhat value close to 1.  This suggests that the chains have converged to the target distribution.

@@

@fig-stanareyouokay-b-2 is a Rhat plot for the negative binomial regression model. It shows an Rhat value close to 1.  This again suggests that the chains have converged to the target distribution in the negative regression model.

>>>>

---
title: "Decoding Mortality: Trends and Causes in Canada between 2001-2022"
subtitle: "A Bayesian Statistical Analysis Unveiling the Predominance of Malignant Neoplasms and Heart Diseases"
author: 
  - Yuchao Niu
thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
date: yesterday
date-format: long
abstract: "The study of mortality offers key public health insights into the fundamental causes and trends of death. This research investigates Canada's mortality trends from 2001 to 2022, employing Bayesian Poisson and negative binomial regression models to analyze data from the Canadian Vital Statistics Death Database. It reveals that malignant neoplasms and heart diseases significantly elevate mortality rates, surpassing the effects of unintentional accidents by five times and three times, respectively. Such findings highlight the critical necessity for targeted public health initiatives focusing on the prevention, early detection, and treatment of these conditions."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)
library(gt)
library(arrow)
library(dplyr)
library(broom.mixed)

@@

The original dataset obtained from the Statistics Canada open government portal includes 2,268 observations across 18 variables, offering detailed demographic and geographical information on registered deaths. I aim to focus on five specific variables: 'Reference Period', 'Leading Causes of Death (ICD-10)', 'Characteristics', 'Ranking', and 'Number of Deaths'. The 'Reference Period' denotes the year of death registration (2000-2022), while 'Leading Causes of Death' categorizes the cause according to the International Statistical Classification of Diseases and Related Health Problems (ICD-10). 'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the cause's rank in a given year, respectively. A sample of the cleaned data can be found in @sec-data-details.

R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@Knitr] , ggplot2[@ggplot2], broom[@broom], dplyr[@dplyr] and  have been used in data downloading, cleaning and visualization.

@@

```{r}
#| label: tbl-1
#| tbl-cap: "Annual number of deaths for the top-five causes in 2021, since 2001, for Canada"
#| echo: false
#| warning: false
#| message: false

cleaned_data <- read_parquet("../data/cleaned_data.parquet")

@@

@@

```{r}
#| label: fig-1
#| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
#| echo: false
#| warning: false
#| message: false

top_five <- transformed_data |>
  filter(n == 22) |>
  slice_max(order_by = desc(Ranking), n = 5)

top_five_causes <- top_five |>
  pull(Cause)

analysis_data <- cleaned_data |>
  filter(Cause %in% top_five_causes)

analysis_data |>
  ggplot(aes(x = Year, y = Death, color = Cause)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Year", y = "Annual number of deaths in Canada") +
  facet_wrap(vars(Cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")

```{r}
#| label: fig-2
#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
#| echo: false
#| warning: false
#| message: false

# cleaned_data |>
#   filter(Year == 2021) |>
#   mutate(Percentage = Death / sum(Death) * 100) |>
#   ggplot(top_five, aes(x="", y= Death, fill= Cause)) +
#   geom_bar(width = 1, stat = "identity") +
#   coord_polar("y") +
#   theme_void() +
#   labs(fill="Cause of Death")

@@

\begin{align} 
y_i&\sim \text{Poisson}(\lambda_i) \quad \\
\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \\
\beta_0 &\sim \text{Normal}(0, 2.5) \quad \\
\beta_1 &\sim \text{Normal}(0, 2.5) \quad 
\end{align}

@@

\begin{align} 
y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad  \\
\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \\
\beta_0 &\sim \text{Normal}(0, 2.5) \quad \\
\beta_1 &\sim \text{Normal}(0, 2.5) \quad \\
\phi &\sim \text{Gamma}(2, 0.1) \quad 
\end{align}

@@

```{r}
#| label: tbl-2
#| tbl-cap: "Summary Statistics of the number of yearly deaths, by cause, in Canada"
#| echo: false

@@

@tbl-3 presents coefficients from both Poisson and negative binomial regression models for various causes of death. The intercept, representing the baseline of Accidents (unintentional injuries), indicates the log count of deaths when no other specific cause is considered. Coefficients for causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of the heart, and malignant neoplasms reflect log-relative differences in death counts relative to Accidents. Positive coefficients indicate a positive association with the number of deaths.

```{r}
#|message= false
#|warning= false
#|include= false

@@

tidy_poisson <- tidy(canada_death_poisson)
tidy_negbinomial <- tidy(canada_death_negbinomial)

coef_short_names_corrected <- c(
  "(Intercept)" = "(Intercept)",
  "CauseCerebrovascular diseases" = "Cerebrovascular diseases",
  "CauseChronic lower respiratory diseases" = "Chronic lower respiratory diseases",
  "CauseDiseases of heart" = "Diseases of heart",
  "CauseMalignant neoplasms" = "Malignant neoplasms",
  "CauseCOVID-19" = "COVID-19"  # Ensure this matches the term exactly as it appears in your model output
)

# Apply the corrected mapping to rename terms for the Poisson model
tidy_poisson$term <- coef_short_names_corrected[tidy_poisson$term]
# Handling any terms that did not match by retaining their original names
tidy_poisson$term[is.na(tidy_poisson$term)] <- names(coef_short_names_corrected)[is.na(tidy_poisson$term)]

# Apply the corrected mapping to rename terms for the Negative Binomial model
tidy_negbinomial$term <- coef_short_names_corrected[tidy_negbinomial$term]
# Handling any terms that did not match by retaining their original names
tidy_negbinomial$term[is.na(tidy_negbinomial$term)] <- names(coef_short_names_corrected)[is.na(tidy_negbinomial$term)]

# Combine the data for comparison
comparison <- bind_rows(
  mutate(tidy_poisson, model = "Poisson"),
  mutate(tidy_negbinomial, model = "Negative Binomial")
)

# Assuming 'comparison' is your final data frame from previous steps
comparison_table <- comparison %>%
  gt() %>%
  tab_header(
    title = "Model Comparison: Poisson vs. Negative Binomial"
  ) %>%
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Standard Error",
    model = "Model"
  ) %>%
  fmt_number(
    columns = c(estimate, std.error),
    decimals = 3
  ) %>%
  tab_spanner(
    label = "Statistics",
    columns = c(estimate, std.error)
  ) %>%
  tab_style(
    style = cell_text(align = 'center'),
    locations = cells_body(columns = everything())
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    table.font.size = 12
  )
```

```{r}
#| label: tbl-3
#| tbl-cap: "Modeling the most prevalent cause of deaths in Canada, 2001-2022"
#| echo: false
#| warning: false
#| message: false

comparison_table
```

@@

Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts. The negative binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794. The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-4. Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.

```{r}
#| label: tbl-4
#| tbl-cap: "Cross validation comparision between Poisson model and negative binomial model "
#| echo: false
#| warning: false
#| message: false

@@

(loo_compare(poisson, neg_binomial))

@@

While the Poisson and negative binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors. Additionally, we only considered one variable the leading cause of death, without considering rarer causes or potential confounders like socioeconomic factors. This constrains the ability of the model to capture the nuances of mortality and its drivers.

@@

The raw dataset was downloaded from Statistics Canada open data portal and cleaned by selecting the variables of interest "Year", "Causes of Death", "Characteristics" and "Value". The "Characteristics" variable is converted into a new variable "Death" indicating the number of death attributed to the cause and a new "Rank" variable showing the rank of the cause in a given year. @tbl-5 presents a sample of the cleaned dataset.

```{r}
#| label: tbl-5
#| tbl-cap: "A sample of the cleaned dataset of number of death attributed to causes of death in Canada, 2001-2022"
#| echo: false
#| warning: false
#| message: false
sampled_data <- sample_n(cleaned_data, 5) %>%
  kable()
sampled_data
```

@@

In @fig-ppcheck-a-1 we implement a posterior predictive check for the Poisson regression model described in @sec-result. This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align reasonably well with the actual data (represented by the dark blue line) with minor misalignment. This suggests the model represent the data well.

In @fig-ppcheck-a-2 we implement a posterior predictive check for the negative binomial regression model described in @sec-result This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align generally well with the actual data (represented by the dark blue line). This suggests the model represents the data well.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck-a
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
#| fig-subcap: ["Poisson model", "Negative binomial model"]

@@

## Diagnostics

@fig-stanareyouokay-a-1 is a trace plot for the Poisson regression model. It shows a horizontal, dense band of samples without any systematic patterns, drifts, or long periods of stagnation. This pattern suggests that the chain is mixing well and sampling efficiently from the posterior distribution. This suggests the Poisson regression model is suitable.

@fig-stanareyouokay-a-2 is a Rhat plot. It shows an Rhat value falling in the range 1-1.1.  This suggests that the chains have converged to the target distribution in the Poisson model.

@@

@fig-stanareyouokay-b-2 is a Rhat plot for the negative binomial regression model. It shows an Rhat value close to 1.  This again suggests that the chains have converged to the target distribution in the negative regression model. The Rhat value is closer to 1 compared to in the Poisson model showing that the negative binomial regression model fits the data better.
